---
title: "tutorial_data_intro"
author: "Mir Henglin"
date: "6/25/2019"
output: html_document
---

To construct out plots in R, we will be using the `ggplot2` package. To perform data manipulation, we will be using the `dplyr` package.

```{r library, warning=FALSE, message=FALSE}
library(dplyr)
library(ggplot2)
```

```{r import, include = FALSE, message=FALSE, warning=FALSE}
plot_data <-
  readRDS('../../../data/tutorial_plot_data.rds') 
```

## Data Format

The plots we will make will summarize the results of multiple models at the same time. In order to plot those results in `ggplot2`, they must be properly formatted. Specifically, the data must be organized in a `data.frame` with columns indicating:

* The model that the results a row correspond to
* The dependent variable term
* The P-value
* The effect size estimate

Other columns can be included in the data, but for the purpose of this tutorial we will focus on the four necessary columns. 
 
### Tutorial Dataset

The dataset we will be using contains regression coefficents and P-values for models evaluating the relationship between the measured levels of metabolites found in the blood and biological measures of interest. A small portion of that dataset is shown below.

```{r}
plot_data
```

In this dataset, `estimate` and `p.value` indicate the effect size estimate and the P-value of that estimate, `term` indicates the ID of the metabolite, and`response` indicates the biological measure of interest. For example, if `response` = Body Mass Index and `term` = mzid_396.271758_6.3118, that row in the dataset corresponds to the model;

$$Body Mass Index = \beta_0 + \beta_1 * mzid\_396.271758\_6.3118 + X\beta + \epsilon$$

Where $X\beta$ corresponds to control variables included in the model and $\epsilon$ is the error term. `estimate` is the estimated value of $\beta_1$ and `p.value` is the corresponding P-value.
