---
title: "Plotting Rainplots"
author: "Mir Henglin"
date: "July 09, 2019"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.height = 10, fig.width = 10, fig.align = "center")
```

```{r data_intro, child = 'sections/tdi4.Rmd'}
```

```{r nl10pv, child = 'sections/nl10pv.Rmd'}
```

## Plotting

A basic rainplot can be constructed in only two lines of `ggplot2` code!

```{r}

rainplot <-
  ggplot(plot_data) +
  geom_point(aes(x = response, y = term, colour = estimate, size = p.value)) 

rainplot
```

This is a good start, but we want to clean up layout and presentation. We can do this by creating a custom `ggplot2` theme and adjusting scales and layout. One thing to ensure is to represent P-values (the size of the plotted points) by area instead of radius. When comparing two points of different size, humans perceive the area of points, not their radius, when comparing them. Thus a value that is twice another should have twice as much area, not double the radius. This is ensured by using `scale_size_area`.

```{r}

thm <- 
  # Good starting theme + set text size
  theme_light(base_size = 18) +
  theme(
    # Remove axis ticks and titles
    axis.title.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.title.y = element_blank(),
    axis.ticks.y = element_blank(),
    
    # Remove gridlines and boxes
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.line = element_blank(),
    legend.key = element_blank(),
    
    # White backgrounds
    panel.background = element_rect(fill = 'white'),
    plot.background = element_rect(fill = 'white'),
    legend.background = element_rect(fill = 'white'),
    
    # Angle text
    axis.text.x.top  = element_text(angle = 45, hjust = 0)
  ) 

rainplot <- 
  rainplot + 
  thm + 
  # Move x-axis label to top of plot
  scale_x_discrete(position = 'top') +
  scale_size_area(expression(paste(-log[10]('P-value')))) + 
  scale_color_continuous('Effect Size Estimate')

rainplot


```

To make the presentation of the effect size estimates clearer, we we use a diverging red-blue palette. In order to guarantee a symmetric palette, we set the palette limits based on our data.

```{r}

palette <-
  # Blue
  c("#053061",
    "#313695",
    "#4575b4",
    "#74add1",
    "#abd9e9",
    "#e0f3f8",
    "#fee090",
    "#fdae61",
    "#f46d43",
    "#d73027",
    "#a50026",
    '#67001f')
# Red

# Calculate symmetric limits based on most extreme value
max_abs_estimate <- max(abs(plot_data$estimate))

max_lim <- max_abs_estimate
min_lim = -1 * max_lim

```

```{r}
rainplot <- rainplot +
  scale_color_gradientn(
    'Effect Size Estimate',
    colors = palette,
    limits = c(min_lim, max_lim),
    breaks = c(min_lim,  min_lim / 2, 0 , max_lim/2, max_lim)
  )

rainplot


```

Another step to improve presentation is to increase the maximum size of each point There will be a bit of trial and error here; if the size threshold is too large, the points will overlap.

```{r}
rainplot + 
  scale_size_area(expression(paste(-log[10]('P-value'))), max_size = 12)
```

## Additional Plot Adjustments

### Outlines

The points on rainplots can be outlined in a color different from the color of the point. To get such a shape, we add the argument `shape = 21` to `geom_point`. Note that when we do this, the color of the point changes from `color` to `fill`. Here, we leave the outlines as the default black. 

```{r}

rainplot <-
  ggplot(plot_data) +
  geom_point(aes(x = response, y = term, fill = estimate, size = p.value), 
             shape = 21) + 
  scale_fill_gradientn(
    'Effect Size Estimate',
    colors = palette,
    limits = c(min_lim, max_lim),
    breaks = c(min_lim, min_lim / 2, 0 , max_lim / 2, max_lim)
  ) +
  scale_size_area(expression(paste(-log[10]('P-value'))), max_size = 12) +
  scale_x_discrete(position = 'top') +
  thm

rainplot
```

### P-value Thresholding

When a few P-values are much smaller than the majority of the data, a rainplot loses size resolution in the range where most of the data lies. One possible solution is to set all P-values above some ceiling, here chosen to be 15, to the value of the ceiling. The threshold can be set at a level where one considers all P-values more extreme than the threshold to be 'of interest'.

```{r}

plot_data_thresholded <- 
  plot_data %>% 
  mutate(p.value = ifelse(p.value > 15, 15, p.value))

rainplot <-
  # Use the thresholded data
  ggplot(plot_data_thresholded) +
  geom_point(aes(x = response, y = term, colour = estimate, size = p.value)) +
  scale_x_discrete(position = 'top') +
  scale_color_gradientn(
    'Effect Size Estimate',
    colors = palette,
    limits = c(min_lim, max_lim),
    breaks = c(min_lim, min_lim / 2, 0 , max_lim / 2, max_lim)
  ) +
  # Set the legend breaks and labels to account for the thresholding
  scale_size_area(
    expression(paste(-log[10]('P-value'))), 
    max_size = 12,
    breaks = c(5, 10, 15), 
    labels = c('5', '10', '>=15')) +
  thm

rainplot
```


### Ordering by P-Value

To make it easier to identify the metabolites that had small P-values in multiple models, we will convert the `term` column into an 'ordered factor' ordered by the average P-value across all models. This will put metabolites with small P-values in multiples models at the top of the plot, and metabolites with large P-values in multiple models at the bottom of the plot.

```{r}

# Calculate mean P-value for each metabolite
mpv <-
  plot_data %>% 
  group_by(term) %>% 
  summarise(mean_pv = mean(p.value))

# Order metabolites by average p-value
term_order <-
  mpv %>% 
  arrange(mean_pv) %>% 
  pull(term)

# Convert term to a factor, ordered by `term_order`
plot_data_pvo <- 
  plot_data %>% 
  mutate(term = factor(term, levels = term_order))
```

```{r}
rainplot <-
  # Use the data with the term column ordered by mean P-value
  ggplot(plot_data_pvo) +
  geom_point(aes(x = response, y = term, colour = estimate, size = p.value)) +
  scale_color_gradientn(
    'Effect Size Estimate',
    colors = palette,
    limits = c(min_lim, max_lim),
    breaks = c(min_lim, min_lim / 2, 0 , max_lim / 2, max_lim)
  ) +
  scale_size_area(expression(paste(-log[10]('P-value'))), max_size = 12) +
  scale_x_discrete(position = 'top') +
  thm

rainplot
```

### Ordering by Cluster

rainplots can be ordered by cluster such that similar dependent variable terms are plotted next to one another. We will be using the `hculst` function to cluster the results by effect size estimate. The `term` variable will be converted into an ordered factor, such that clustered terms are plotted next to one another. In order to cluster the data, we will need to reshape it using the spread function from the `tidyr` package. 

```{r, warning=FALSE, message=FALSE}
library(tidyr)
```

```{r}
# Convert to matrix and reshape for clustering.
cluster_data <-
  plot_data %>% 
  select(response, term, estimate) %>% 
  spread(response, estimate)

rnms <- 
  cluster_data$term

cluster_data <-
  cluster_data %>% 
  select(-term) %>% 
  as.matrix()

rownames(cluster_data) <- rnms

# Quick peek to see structure of data
cluster_data[1:5, 1:5]
```

```{r}
# cluster dependent variable terms
clust <- hclust(dist(cluster_data), method = 'ward.D2')
```

```{r}

# `clust$order` orders `term` into clusters
term_order <-
  clust$labels[clust$order]

# Convert term to a factor, ordered by `term_order`
plot_data_clo <- 
  plot_data %>% 
  mutate(term = factor(term, levels = term_order))
```

```{r}
rainplot <-
  # Use cluter ordered data
  ggplot(plot_data_clo) +
  geom_point(aes(x = response, y = term, colour = estimate, size = p.value)) +
  scale_color_gradientn(
    'Effect Size Estimate',
    colors = palette,
    limits = c(min_lim, max_lim),
    breaks = c(min_lim, min_lim / 2, 0 , max_lim / 2, max_lim)
  ) +
  scale_size_area(expression(paste(-log[10]('P-value'))), max_size = 12) +
  scale_x_discrete(position = 'top') +
  thm

rainplot
```

```{r nl10pv, child = 'sections/sbsp.Rmd'}
```

### Adding dendrograms

Dendrograms can be added to cluster-ordered `ggplot2` plots using the `ggdendro` package. 

```{r}
library(ggdendro)
```

```{r}
# Extract dendrogram data from previous cluster results
dendro_dat <- segment(dendro_data(clust))
```

A basic dendrogram can be plotted quite easily.

```{r}
dendro <-
  ggplot(dendro_dat) +
  geom_segment(aes(x = x, y = y, xend=xend, yend=yend), colour = 'black') 

dendro
```

The default dendrogram points down. To put the dendrogram on the left of our plot, we want it to point to the right. We can do this by switching the x and y coordinates.

```{r}
dendro <-
  ggplot(dendro_dat) +
  geom_segment(aes(x = -y, y = x, xend=-yend, yend=xend), colour = 'black')

dendro
```

Before plotting our dendrogram and rainplot side-by-side, we need to first create a version of our dendrogram with the same scale and layout as our rainplot

```{r}

# Taking only the column with the longest label maintains alignment while
# simplifying plot layout
x_labels <- 
  plot_data_clo$response %>% 
  unique()

longest_x_label <-
  x_labels[[which.max(nchar(x_labels))]]

longest_x_label_data <-
  plot_data %>% 
  filter(response == longest_x_label)

```

```{r}
# Align the top of the dendrogram with the x-axis label
max_dendro <-
  max(abs(c(dendro_dat$y, dendro_dat$yend)))

offset <- max_dendro + 1
```


```{r}

dendro <-
  # Empty ggplot with same layout and scale as rainplot
  ggplot() +
  geom_blank(aes(x = response, y = term), data = longest_x_label_data) + 
  thm +
  # 'expand' controls whitespace around the dendrogram. The non-zero argument
  # may need to be increasesed if the line thickness of the dendrogram is
  # increased to make sure the entire dendrogram is plotted
  scale_x_discrete(position = 'top', expand = c(0, 0.02, 0, 0)) +
  # Draw dendrogram
  geom_segment(aes(x = (-y + offset), y = x, xend = (-yend + offset), yend=xend),
               colour = 'black', 
               data = dendro_dat) +
  # Mave invisble unnessecary plot layout.
  theme(legend.position = 'none',
        axis.text.y = element_blank(),
        axis.text.x = element_text(colour = 'white'),
        panel.grid = element_blank(),
        panel.border = element_rect(fill = NA, colour = 'white')
  )  



grid.arrange(dendro, rainplot, ncol = 2, widths = c(3, 15)) 
```



